### 2021-10-29 papers/thesis/resubmitted/ch2-spherical-harmonics/spherical-harmonics.tex
Added bstodo notes to indicate changes

To introduce ourselves to the world of multidimensional orthogonal polynomials for solving PDEs on the sphere, we can naturally choose to look at the famous spherical harmonics. Our aim here is to express the spherical harmonics as polynomials in three variables $x$, $y$, and $z$ to evaluate functions and solve PDEs on the whole sphere. More precisely, we desire the solution to partial differential equations on the domain \nomenclature[Omega]{$\Omega$}{The domain of interest}
\bseqn{
	\Omega := \bigS^2 = \bsset{(x,y,z) \in \R^3}{x^2 + y^2 = \rho(z)^2},
}
where 
\bseqn{
	\rho(z) := \sqrt{1-z^2}.
}
While it may seem somewhat odd, it should hopefully become apparent why using $\rho$ here is a practical way to define our domain.

@@

Before we proceed further, let us build up to our definition of the spherical harmonics. We first need to introduce a few classical orthogonal polynomials. Orthogonal polynomials are defined as such by the condition that each polynomial is orthogonal to each other, with respect to a given inner product, and ordered by degree. In particular, they are orthogonal with respect to all lower degree polynomials.

On the unit interval, $[-1,1]$, we note that there is a hierarchy of OPs in the sense that \cite[Table 18.3.1, (18.9.15)]{DLMF}: \nomenclature[Pnab]{$P^{(a,b)}_n$}{The degree $n$ Jacobi polynomial (context determines normalized or not) unless otherwise stated}
\bseqn{
	& \ddx{x} P^{(a,b)}_l (x) = \half \: (l + a + b + 1) \: P^{(a+1,b+1)}_{l-1}(x), \\
	\implies & \dmdxm{x}{m} P_l(x) = {(l+m)! \over 2^m \: l!} \: P^{(m,m)}_{l-m}(x),
}
where $P^{(a,b)}_l (x)$ is the degree $l$ \textit{Jacobi polynomial} with $a, b > -1$, and \nomenclature[Pl]{$P_l$}{The degree $l$ Legendre polynomial} $P_l(x) := P^{(0,0)}_l (x)$ is simply the \textit{Legendre polynomial} of degree $l$. Jacobi polynomials are orthogonal with respect to the weight $w(x) = (1-x)^a(1+x)^b$; that is for $l, l' \in \No,$ \nomenclature[Kronecker delta]{$\delta_{n, n'}$}{The Kronecker delta function; $1$ if $n=n'$, else $0$}
\bseqn{
	\int_{-1}^1 {P^{(a,b)}_l(x)}^2 \:(1-x)^a(1+x)^b \: \D x &=: \omega_{P,l}^{(a,b)}, \\
	\int_{-1}^1 P^{(a,b)}_l(x) \: P^{(a,b)}_{l'}(x) \:(1-x)^a(1+x)^b \: \D x &= \omega_{P,l}^{(a,b)} \: \delta_{l, l'},
}
where $\delta_{l, l'}$ is the regular Kronecker delta function. For the Jacobi polynomials,
\bseqn{
	 \omega_{P,l}^{(a,b)} = {2^{a + b + 1} \; \Gamma(l + a + 1) \; \Gamma(l + b + 1) \over (2l + a + b + 1) \; \Gamma(l + a + b + 1) \; \Gamma(l + 1)},
}
where $\Gamma(x)$ is the regular Gamma function \cite[(5.2.1)]{DLMF} (with $\Gamma(n) = (n-1)!$ for $n \in \N$). Legendre polynomials are special cases of \textit{Legendre functions} when $l$ is a non-negative integer. Legendre functions are a class of univariate functions that are one set of solutions to the differential equation
\bseqn{
	(1 - x^2) \dmdxm{x}{2}y - 2x \ddx{x} y + l(l+1)y = 0, \quad x \in [-1,1]
}
which is helpfully known as the \textit{Legendre equation} \cite[(14.2.1)]{DLMF}. Further, the \textit{associated Legendre polynomials} are a set of polynomials orthogonal with respect to unit weight on the unit interval, and can be defined as the $m^{\text{th}}$ derivative of a Legendre polynomial \cite[p.5]{belousov1962tables}: \nomenclature[Plm]{$\Plm$}{The Associated Legendre polynomial of degree $l$ and order $m$}
\bseqn{
	\Plm(x) &:= (-1)^m (1-x^2)^\frac{m}{2} \dmdxm{x}{m} P_l(x) = (-1)^m \; {(l+m)! \over 2^m \; l!} \; (1-x^2)^\frac{m}{2} \; P^{(m,m)}_{l-m}(x),
}
for $m = 0,1,2,\dots,l$. It is also standard to include a further definition for the case of $m$ negative -- that is we also define:
\bseqn{
	\Plm(x) &:= (-1)^m \; {(l+m)! \over (l-m)!} \; P_l^{\abs{m}}(x) = {(l+m)! \over 2^{\abs{m}} \; l!} \; (1-x^2)^{\abs{m} \over 2} \; \jaclmabs{l}{m}(x),
}
for $m = -l, \dots, -1$.

@@

On that note, we can now write down the spherical harmonics. Let $(x,y,z) \in \Omega$. We will use the standard definition -- that is, the spherical harmonics, orthonormal on the unit sphere, are \cite[(14.30.1)]{DLMF}: \nomenclature[Ylm]{$\Ylm$}{The degree $l$ and order $m$ spherical harmonic}
\bseqnnumber{
	\Ylmfull &:= \Bigg({(2l+1) \: (l-m)! \over 4 \pi \: (l+m)!}\Bigg)^\frac{1}{2} \: \eimtheta \Plm (\cosphi) \nonumber \\
	&= \clm (1 - (\cosphi)^2)^\frac{|m|}{2} \eimtheta P^{(|m|,|m|)}_{l-|m|}(\cosphi) \nonumber \\
	&= \clm P^{(|m|,|m|)}_{l-|m|}(z) \: \rho(z)^{|m|} \: \eimtheta, \label{eqn:shdef}
}
for $0 \le |m| \le l, \, l \in \No$, where \nomenclature[clm]{$\clm$}{The normalising constant for the spherical harmonic $\Ylm$}
\bseqnnumber{
	\clm := \Bigg({(2l+1) \: (l-m)! \over 4\pi \: (l+m)!}\Bigg)^\half \: {(l+m)! \over 2^{\abs{m}} \; l!} \:
		\begin{cases} 
			(-1)^m &\text{if } m \ge 0 \\
			1 &\text{if } m < 0,
		\end{cases} \label{eqn:clmdef}
}
which can be simplified to
\bseqn{
	\clm = \Bigg({(2l+1) \: (l+m)! \: (l-m)! \over \pi }\Bigg)^\half \: {1 \over 2^{\abs{m} + 1} \; l!} \:
		\begin{cases} 
			(-1)^m &\text{if } m \ge 0 \\
			1 &\text{if } m < 0.
		\end{cases}
}
The spherical harmonics as defined are then orthonormal with respect to the inner product with uniform measure \nomenclature[conjuagate]{$a^*$}{The complex conjugate of $a \in \C$}
\bseqn{
	&\iint_{\Omega} \Ylmfull \: {Y_{l'}^{m'}(\varphi, \theta)}^{*} \: \D S \\
	&\quad = \int_0^{2\pi} \int_0^\pi \Ylmfull \: {Y_{l'}^{m'}(\varphi, \theta)}^{*} \: \sinphi \: \D \varphi \: \D \theta \\
	&\quad = 2\pi \: \delta_{m, m'} \: \clm \: c_{l'}^{m'} \: \int_{-1}^1 P^{(|m|,|m|)}_{l-|m|}(z) \: P^{(|m|,|m|)}_{l'-|m|}(z) \: \rho(z)^{2|m|} \: \D z \\
	&\quad = \delta_{l, l'} \: \delta_{m, m'},
}
where $\alpha^*$ denotes the complex conjugate of \nomenclature[C]{$\C$}{The complex numbers} $\alpha \in \C$. Note how we can express the spherical harmonics $\Ylm$ in terms of $x$, $y$, and $z$ instead of $\varphi, \theta$ by noting that $\rho(z)^{|m|} \eimtheta$ can be expressed in terms of $x$, $y$, and $z$ for any \nomenclature[Z]{$\Z$}{The integers} $m \in \Z$. Indeed, they are polynomials in $x$, $y$, and $z$, which we denote $\Ylm(x,y,z)$. They span all polynomials modulo the vanishing ideal associated by the zero set of $x^2 + y^2 + z^2 - 1$ in \nomenclature[R]{$\R$}{The real numbers} \nomenclature[Rd]{$\R^d$}{The d-dimensional coordinate space over the real numbers} $\R^3$.

@@

	The first two results are simply consequences of the relationships established for incrementing and decrementing the parameters of Jacobi polynomials detailed in \cite[(18.9.5), (18.9.6)]{DLMF}, along with the symmetry relation that $P_{n}^{(\alpha, \beta)}(-z) \equiv (-1)^n P_{n}^{(\beta, \alpha)}(z)$. The final one is obtained by rearranging the three-term recurrence in \cite[(18.9.1), (18.9.2)]{DLMF}

@@

For each $l = 0,1,2\dots$ let $\Dlt$ be any matrix that is a left inverse of $A_l$, i.e., such that $\Dlt A_l = \idmat{2l+3}$. Multiplying our system by the preconditioner matrix that is given by the block diagonal matrix of the $\Dlt$'s, we obtain a lower triangular system \cite[\S3.2]{dunkl2014orthogonal}, which can be expanded to obtain the three-term recurrence, for $l = 0,1,2,\dots$:
\bseqn{
	\begin{cases}
		\SHvec_{-1}(x,y,z) := 0, \\
		\SHvec_{0}(x,y,z) := Y_0, \\
		\SHvec_{l+1}(x,y,z) = -\Dlt (B_l-G_l(x,y,z)) \SHvec_l(x,y,z) - \Dlt C_l  \,\SHvec_{l-1}(x,y,z).
	\end{cases}
}
Since the above holds for any $\Dlt$ that is a left inverse of $A_l$, we are free to choose the $\Dlt$ matrices. In order to attempt to maximise sparsity, we define them in the following way\footnote{There may be a more computationally efficient way of defining these matrices, but in practice we only desire the output of the action of $\Dlt C_l$ and $\Dlt (B_l-G_l(x,y,z))$, and so minimising the resulting number of operations is the goal. The choices of the $\Dlt$ matrices we make in this thesis are designed to do this, but whether there are more computationally efficient ways of defining them is an open question.}. For $l \in \N$, we can set
\bseqnnumber{
	\Dlt = 
		\begin{pmatrix}
			& & & & \vec{\eta}^\top_{0} \\
			0 & & & 0 & & & {1 \over \gamma_{l,-l,2}} & \\
			& \ddots & & & \ddots & & & \ddots \\
			& & 0 & & & 0 & & & {1 \over \gamma_{l,l,2}} \\
			& & & & \vec{\eta}^\top_{1} 
		\end{pmatrix} \in \R^{(2l+3)\times3(2l+1)}, \label{eqn:Dltdef}
}
where $\vec{\eta}_{0}, \vec{\eta}_{1} \in \R^{3(2l+1)}$ are vectors with entries given by 
\bseqn{
	\big(\vec{\eta}_{0}\big)_j &= 
		\begin{cases}
			{1 \over \alpha_{l,-l,3}} & j = 1 \\
			{- \; \alpha_{l,-l,4} \over \alpha_{l,-l,3} \: \gamma_{l,1-l,2}} & j = 2(2l+1) + 2 \\
			0 &\text{otherwise},
		\end{cases} \\
	\big(\vec{\eta}_{1}\big)_j &= 
		\begin{cases}
			{1 \over \alpha_{l,l,4}} & j = 2l+1 \\
			{- \; \alpha_{l,l,3} \over \alpha_{l,l,4} \: \gamma_{l,l-1,2}} & j = 3(2l+1) - 1 \\
			0 &\text{otherwise}.
		\end{cases}.
}
For $l = 0$ we can simply set
\bseqnnumber{
	D^T_0 = 
		\begin{pmatrix}
			{\beta_{0,0,4} \over \alpha_{0,0,3} \beta_{0,0,4} - \alpha_{0,0,4} \beta_{0,0,3}} & {- \alpha_{0,0,4} \over \alpha_{0,0,3} \beta_{0,0,4} - \alpha_{0,0,4} \beta_{0,0,3}} & 0 \\
			0 & 0 & {1 \over \gamma_{0,0,2}} \\
			{\beta_{0,0,3} \over \alpha_{0,0,4} \beta_{0,0,3} - \alpha_{0,0,3} \beta_{0,0,4}} & {- \alpha_{0,0,3} \over \alpha_{0,0,4} \beta_{0,0,3} - \alpha_{0,0,3} \beta_{0,0,4}} & 0
		\end{pmatrix}. \label{eqn:Dzerotdef}
}
It will be convenient for us to give a formal name for these coefficient matrices above for a family of multidimensional orthogonal polynomials.
\begin{definition}\label{def:clenshawmats}
	The matrices $-\Dlt (B_l-G_l(x,y,z))$, $\Dlt C_l$ for $l \in \No$ defined via \bsrefeqns{eqn:clenshawmats1}{eqn:Dzerotdef} are called the recurrence coefficient matrices for a given family of multidimensional orthogonal polynomials.
\end{definition}

@@

In coefficient space, we wish to work only with vectors of coefficients for the expansion of a function, to which we can apply operator matrices to that represent differential or other operations. Naturally, we of course need a way to obtain the coefficients $f_{l,m}$. The coefficients can be calculated via the integral
\bseqn{
	f_{l,m} = \iint_\Omega f(\costheta \sinphi, \sintheta \sinphi, \cosphi) \: \Ylmfull^* \: \sinphi \: \D \varphi \: \D \theta,
}
using the orthonormality of the spherical harmonics. As discussed in \bsrefchapter{CHAPTER:intro}, methods to calculate these coefficients exist known as spectral transforms, and are well established for the spherical harmonics (see e.g. \cite{mohlenkamp1999fast, suda2002fast}). Additionally, packages are available for public use that implement efficient spherical harmonics transforms, for example the FastTransforms package \cite{slevinsky2019fast}.

@@

Let $\vec{v}$ be the coefficients of the expansion of $v$ up to order $N$. The operator $V$ would then be the result of the following \textit{operator Clenshaw algorithm}:
\bseqn{
	\quad &\text{1) } \text{Set } \vec{\xi}_{N+2} = \vec{0}, \: \vec{\xi}_{N+1} = \vec{0} \\
	\quad &\text{2) } \text{For } n = N:-1:0 \\
	\quad & \quad \quad \quad \text{set } \vec{\xi}_{n}^T = \vec{v}_n^T - \vec{\xi}_{n+1}^T D^T_n (B_n - G_n(J_x,J_y,J_z)) - \vec{\xi}_{n+2}^T D^T_{n+1}C_{n+1} \\
	\quad &\text{3) } \text{Output: } V = V(x,y,z) \approx \xi_0 Y_0
}
where at each iteration, $\vec{\xi}_n$ is a vector of matrices\footnote{It may help to think of this symbolically. At each iteration, the entries of $\vec{\xi}_n$ are polynomials in $J_x$, $J_y$ and $J_z$. On the final step, $\vec{\xi}_{0} = \xi_{0}$ is hence some polynomial expression in $J_x$, $J_y$ and $J_z$, which we can then evaluate by inserting the Jacobi matrices.} (note that we are abusing notation here a bit, however it is the simplest way to present the algorithm without introducing yet more matrices!).

@@

The Jacobi matrices are then defined according to the matrices \nomenclature[JT]{$\tangentjacobi_x, \tangentjacobi_y, \tangentjacobi_z$}{The Jacobi matrices for the tangent bundle VSH basis} $\tangentjacobi_x, \tangentjacobi_y, \tangentjacobi_z$ satisfying

@@

The framework we have outlined for the spherical harmonics and the vector spherical harmonics allows us to easily and explicitly derive sparse operator matrices for partial differential operators such as the spherical gradient, divergence and Laplace--Beltrami, as well as other key operators found in examples such as the linearised shallow water equations. These operators will take coefficients of a function's expansion in either the SH or VSH basis, to coefficients in either basis.

@@

\begin{theorem}
	The operator matrices defined in \bsrefdef{def:SHoperators} are diagonal, and are given by:
\bseqn{
	\calG &= I_{(N+1)^2}, \quad
	\tilde \calG = 
		\begin{pmatrix}
			1 & & & & \\
			& 0 & & & \\
			& & 1 & & & & \\
			& & & 0 & & & \\
			& & & & \ddots & & \\
			& & & & & 1 & \\
			& & & & & & 0
		\end{pmatrix}  \in \R^{2(N+1)^2 \times 2(N+1)^2}, \\
	\calD &\equiv \calL := 
		\begin{pmatrix}
			\calL_0 & & \\
			& \ddots & \\
			& & \calL_N
		\end{pmatrix} \in \R^{(N+1)^2 \times (N+1)^2}, \\
	\tilde \calD &= 
		\begin{pmatrix}
			\tilde \calD_0 & & \\
			& \ddots & \\
			& & \tilde \calD_N
		\end{pmatrix} \in \R^{2(N+1)^2 \times 2(N+1)^2},
}
where
\bseqn{
	\calL_l := -l(l+1) \: I_{2l+1}, \quad
	\tilde \calD_l := \begin{pmatrix}
			-l(l+1) & & & & \\
			& 0 & & & \\
			& & \ddots & & \\
			& & & -l(l+1) & \\
			& & & & 0
		\end{pmatrix} \in \R^{2(2l+1) \times 2(2l+1)}.
}.
\end{theorem}

>>>>

To introduce ourselves to the world of multidimensional orthogonal polynomials for solving PDEs on the sphere, we can naturally choose to look at the famous spherical harmonics. Our aim here is to express the spherical harmonics as polynomials in three variables $x$, $y$, and $z$ to evaluate functions and solve PDEs on the whole sphere. More precisely, we desire the solution to partial differential equations on the domain \nomenclature[Omega]{$\Omega$}{The domain of interest} \bstodo{Added that this domain is $\bigS^2$}
\bseqn{
	\Omega := \bigS^2 = \bsset{(x,y,z) \in \R^3}{x^2 + y^2 = \rho(z)^2},
}
where 
\bseqn{
	\rho(z) := \sqrt{1-z^2}.
}
While it may seem somewhat odd, it should hopefully become apparent why using $\rho$ here is a practical way to define our domain.

@@

Before we proceed further, let us build up to our definition of the spherical harmonics. We first need to introduce a few classical orthogonal polynomials. Orthogonal polynomials are defined as such by the condition that each polynomial is orthogonal to each other, with respect to a given inner product, and ordered by degree.\bstodo{added that they are degree ordered} In particular, they are orthogonal with respect to all lower degree polynomials.

On the unit interval, $[-1,1]$, we note that there is a hierarchy of OPs in the sense that \cite[Table 18.3.1, (18.9.15)]{DLMF}: \nomenclature[Pnab]{$P^{(a,b)}_n$}{The degree $n$ Jacobi polynomial (context determines normalized or not) unless otherwise stated}
\bseqn{
	& \ddx{x} P^{(a,b)}_l (x) = \half \: (l + a + b + 1) \: P^{(a+1,b+1)}_{l-1}(x), \\
	\implies & \dmdxm{x}{m} P_l(x) = {(l+m)! \over 2^m \: l!} \: P^{(m,m)}_{l-m}(x),
}
where $P^{(a,b)}_l (x)$ is the degree $l$ \textit{Jacobi polynomial} with $a, b > -1$ \bstodo{added the a,b constraint}, and \nomenclature[Pl]{$P_l$}{The degree $l$ Legendre polynomial} $P_l(x) := P^{(0,0)}_l (x)$ is simply the \textit{Legendre polynomial} of degree $l$. Jacobi polynomials are orthogonal with respect to the weight $w(x) = (1-x)^a(1+x)^b$; that is for $l, l' \in \No,$ \nomenclature[Kronecker delta]{$\delta_{n, n'}$}{The Kronecker delta function; $1$ if $n=n'$, else $0$}
\bseqn{
	\int_{-1}^1 {P^{(a,b)}_l(x)}^2 \:(1-x)^a(1+x)^b \: \D x &=: \omega_{P,l}^{(a,b)}, \\
	\int_{-1}^1 P^{(a,b)}_l(x) \: P^{(a,b)}_{l'}(x) \:(1-x)^a(1+x)^b \: \D x &= \omega_{P,l}^{(a,b)} \: \delta_{l, l'},
}
where $\delta_{l, l'}$ is the regular Kronecker delta function.\bstodo{clarified that this delta is the Kronecker} For the Jacobi polynomials,
\bseqn{
	 \omega_{P,l}^{(a,b)} = {2^{a + b + 1} \; \Gamma(l + a + 1) \; \Gamma(l + b + 1) \over (2l + a + b + 1) \; \Gamma(l + a + b + 1) \; \Gamma(l + 1)},
}
where $\Gamma(x)$ is the regular Gamma function \cite[(5.2.1)]{DLMF} (with $\Gamma(n) = (n-1)!$ for $n \in \N$). Legendre polynomials are special cases of \textit{Legendre functions} when $l$ is a non-negative integer. Legendre functions are a class of univariate functions that are one set of solutions to the differential equation \bstodo{clarified that they are one set of solutions to the following eqn. Added the x domain.}
\bseqn{
	(1 - x^2) \dmdxm{x}{2}y - 2x \ddx{x} y + l(l+1)y = 0, \quad x \in [-1,1]
}
which is helpfully known as the \textit{Legendre equation} \cite[(14.2.1)]{DLMF}. Further, the \textit{associated Legendre polynomials} are a set of polynomials orthogonal with respect to unit weight on the unit interval, and can be defined as the $m^{\text{th}}$ derivative of a Legendre polynomial \cite[p.5]{belousov1962tables}: \nomenclature[Plm]{$\Plm$}{The Associated Legendre polynomial of degree $l$ and order $m$}
\bseqn{
	\Plm(x) &:= (-1)^m (1-x^2)^\frac{m}{2} \dmdxm{x}{m} P_l(x) = (-1)^m \; {(l+m)! \over 2^m \; l!} \; (1-x^2)^\frac{m}{2} \; P^{(m,m)}_{l-m}(x),
}
for $m = 0,1,2,\dots,l$. It is also standard to include a further definition for the case of $m$ negative -- that is we also define:
\bseqn{
	\Plm(x) &:= (-1)^m \; {(l+m)! \over (l-m)!} \; P_l^{\abs{m}}(x) = {(l+m)! \over 2^{\abs{m}} \; l!} \; (1-x^2)^{\abs{m} \over 2} \; \jaclmabs{l}{m}(x),
}
for $m = -l, \dots, -1$.

@@

On that note, we can now write down the spherical harmonics. Let $(x,y,z) \in \Omega$. We will use the standard definition -- that is, the spherical harmonics, orthonormal on the unit sphere, are \cite[(14.30.1)]{DLMF}: \nomenclature[Ylm]{$\Ylm$}{The degree $l$ and order $m$ spherical harmonic}
\bseqnnumber{
	\Ylmfull &:= \Bigg({(2l+1) \: (l-m)! \over 4 \pi \: (l+m)!}\Bigg)^\frac{1}{2} \: \eimtheta \Plm (\cosphi) \nonumber \\
	&= \clm (1 - (\cosphi)^2)^\frac{|m|}{2} \eimtheta P^{(|m|,|m|)}_{l-|m|}(\cosphi) \nonumber \\
	&= \clm P^{(|m|,|m|)}_{l-|m|}(z) \: \rho(z)^{|m|} \: \eimtheta, \label{eqn:shdef}
}
for $0 \le |m| \le l, \, l \in \No$, where \nomenclature[clm]{$\clm$}{The normalising constant for the spherical harmonic $\Ylm$}
\bseqnnumber{
	\clm := \Bigg({(2l+1) \: (l-m)! \over 4\pi \: (l+m)!}\Bigg)^\half \: {(l+m)! \over 2^{\abs{m}} \; l!} \:
		\begin{cases} 
			(-1)^m &\text{if } m \ge 0 \\
			1 &\text{if } m < 0,
		\end{cases} \label{eqn:clmdef}
}
which can be simplified to
\bseqn{
	\clm = \Bigg({(2l+1) \: (l+m)! \: (l-m)! \over \pi }\Bigg)^\half \: {1 \over 2^{\abs{m} + 1} \; l!} \:
		\begin{cases} 
			(-1)^m &\text{if } m \ge 0 \\
			1 &\text{if } m < 0.
		\end{cases}
}
The spherical harmonics as defined are then orthonormal with respect to the inner product with uniform measure \nomenclature[conjuagate]{$a^*$}{The complex conjugate of $a \in \C$} \bstodo{Added that the following is $\iint_{\Omega} \Ylmfull \: {Y_{l'}^{m'}(\varphi, \theta)}^{*} \: \D S$.}
\bseqn{
	&\iint_{\Omega} \Ylmfull \: {Y_{l'}^{m'}(\varphi, \theta)}^{*} \: \D S \\
	&\quad = \int_0^{2\pi} \int_0^\pi \Ylmfull \: {Y_{l'}^{m'}(\varphi, \theta)}^{*} \: \sinphi \: \D \varphi \: \D \theta \\
	&\quad = 2\pi \: \delta_{m, m'} \: \clm \: c_{l'}^{m'} \: \int_{-1}^1 P^{(|m|,|m|)}_{l-|m|}(z) \: P^{(|m|,|m|)}_{l'-|m|}(z) \: \rho(z)^{2|m|} \: \D z \\
	&\quad = \delta_{l, l'} \: \delta_{m, m'},
}
where $\alpha^*$ denotes the complex conjugate of \nomenclature[C]{$\C$}{The complex numbers} $\alpha \in \C$. Note how we can express the spherical harmonics $\Ylm$ in terms of $x$, $y$, and $z$ instead of $\varphi, \theta$ by noting that $\rho(z)^{|m|} \eimtheta$ can be expressed in terms of $x$, $y$, and $z$ for any \nomenclature[Z]{$\Z$}{The integers} $m \in \Z$. Indeed, they are polynomials in $x$, $y$, and $z$, which we denote $\Ylm(x,y,z)$. They span all polynomials modulo the vanishing ideal associated by the zero set of $x^2 + y^2 + z^2 - 1$ in \nomenclature[R]{$\R$}{The real numbers} \nomenclature[Rd]{$\R^d$}{The d-dimensional coordinate space over the real numbers} $\R^3$.

@@

	The first two results are simply consequences of the relationships established for incrementing and decrementing the parameters of Jacobi polynomials detailed in \cite[(18.9.5), (18.9.6)]{DLMF}, along with the symmetry relation that $P_{n}^{(\alpha, \beta)}(-z) \equiv (-1)^n P_{n}^{(\beta, \alpha)}(z)$. The final one is obtained by rearranging the three-term recurrence in \cite[(18.9.1), (18.9.2)]{DLMF}. \bstodo{Reworded this sentence.}

@@

For each $l = 0,1,2\dots$ let $\Dlt$ be any matrix that is a left inverse of $A_l$, i.e., such that $\Dlt A_l = \idmat{2l+3}$. Multiplying our system by the preconditioner matrix that is given by the block diagonal matrix of the $\Dlt$'s, we obtain a lower triangular system \cite[\S3.2]{dunkl2014orthogonal}, which can be expanded to obtain the three-term recurrence, for $l = 0,1,2,\dots$:
\bseqn{
	\begin{cases}
		\SHvec_{-1}(x,y,z) := 0, \\
		\SHvec_{0}(x,y,z) := Y_0, \\
		\SHvec_{l+1}(x,y,z) = -\Dlt (B_l-G_l(x,y,z)) \SHvec_l(x,y,z) - \Dlt C_l  \,\SHvec_{l-1}(x,y,z).
	\end{cases}
}
Since the above holds for any $\Dlt$ that is a left inverse of $A_l$, we are free to choose the $\Dlt$ matrices. In order to attempt to maximise sparsity, we define them in the following way\footnote{There may be a more computationally efficient way of defining these matrices, but in practice we only desire the output of the action of $\Dlt C_l$ and $\Dlt (B_l-G_l(x,y,z))$, and so minimising the resulting number of operations is the goal. The choices of the $\Dlt$ matrices we make in this thesis are designed to do this, but whether there are more computationally efficient ways of defining them is an open question.}.\bstodo{Added footnote to help explain choice of $\Dlt$'s} For $l \in \N$, we can set \bstodo{Tried to reformat the matrix defn to make it clearer.}
\bseqnnumber{
	\Dlt = 
		\begin{pmatrix}
			& & & & \vec{\eta}^\top_{0} \\
			0 & & & 0 & & & {1 \over \gamma_{l,-l,2}} & \\
			& \ddots & & & \ddots & & & \ddots \\
			& & 0 & & & 0 & & & {1 \over \gamma_{l,l,2}} \\
			& & & & \vec{\eta}^\top_{1} 
		\end{pmatrix} \in \R^{(2l+3)\times3(2l+1)}, \label{eqn:Dltdef}
}
where $\vec{\eta}_{0}, \vec{\eta}_{1} \in \R^{3(2l+1)}$ are vectors with entries given by 
\bseqn{
	\big(\vec{\eta}_{0}\big)_j &= 
		\begin{cases}
			{1 \over \alpha_{l,-l,3}} & j = 1 \\
			{- \; \alpha_{l,-l,4} \over \alpha_{l,-l,3} \: \gamma_{l,1-l,2}} & j = 2(2l+1) + 2 \\
			0 &\text{otherwise},
		\end{cases} \\
	\big(\vec{\eta}_{1}\big)_j &= 
		\begin{cases}
			{1 \over \alpha_{l,l,4}} & j = 2l+1 \\
			{- \; \alpha_{l,l,3} \over \alpha_{l,l,4} \: \gamma_{l,l-1,2}} & j = 3(2l+1) - 1 \\
			0 &\text{otherwise}.
		\end{cases}.
}
For $l = 0$ we can simply set
\bseqnnumber{
	D^T_0 = 
		\begin{pmatrix}
			{\beta_{0,0,4} \over \alpha_{0,0,3} \beta_{0,0,4} - \alpha_{0,0,4} \beta_{0,0,3}} & {- \alpha_{0,0,4} \over \alpha_{0,0,3} \beta_{0,0,4} - \alpha_{0,0,4} \beta_{0,0,3}} & 0 \\
			0 & 0 & {1 \over \gamma_{0,0,2}} \\
			{\beta_{0,0,3} \over \alpha_{0,0,4} \beta_{0,0,3} - \alpha_{0,0,3} \beta_{0,0,4}} & {- \alpha_{0,0,3} \over \alpha_{0,0,4} \beta_{0,0,3} - \alpha_{0,0,3} \beta_{0,0,4}} & 0
		\end{pmatrix}. \label{eqn:Dzerotdef}
}
It will be convenient for us to give a formal name for these coefficient matrices above for a family of multidimensional orthogonal polynomials.
\begin{definition}\label{def:clenshawmats}
	The matrices $-\Dlt (B_l-G_l(x,y,z))$, $\Dlt C_l$ for $l \in \No$ defined via \bsrefeqns{eqn:clenshawmats1}{eqn:Dzerotdef} are called the recurrence coefficient matrices for a given family of multidimensional orthogonal polynomials.
\end{definition}

@@

In coefficient space,\bstodo{Changed "spectral space" to "coefficient space".} we wish to work only with vectors of coefficients for the expansion of a function, to which we can apply operator matrices to that represent differential or other operations. Naturally, we of course need a way to obtain the coefficients $f_{l,m}$. The coefficients can be calculated via the integral
\bseqn{
	f_{l,m} = \iint_\Omega f(\costheta \sinphi, \sintheta \sinphi, \cosphi) \: \Ylmfull^* \: \sinphi \: \D \varphi \: \D \theta,
}
using the orthonormality of the spherical harmonics. As discussed in \bsrefchapter{CHAPTER:intro}, methods to calculate these coefficients exist known as spectral transforms, and are well established for the spherical harmonics (see e.g. \cite{mohlenkamp1999fast, suda2002fast}). Additionally, packages are available for public use that implement efficient spherical harmonics transforms, for example the FastTransforms package \cite{slevinsky2019fast}.

@@

Let $\vec{v}$ be the coefficients of the expansion of $v$ up to order $N$. The operator $V$ would then be the result of the following \textit{operator Clenshaw algorithm}:\bstodo{Corrected the typo on the last line of the algorithm.}
\bseqn{
	\quad &\text{1) } \text{Set } \vec{\xi}_{N+2} = \vec{0}, \: \vec{\xi}_{N+1} = \vec{0} \\
	\quad &\text{2) } \text{For } n = N:-1:0 \\
	\quad & \quad \quad \quad \text{set } \vec{\xi}_{n}^T = \vec{v}_n^T - \vec{\xi}_{n+1}^T D^T_n (B_n - G_n(J_x,J_y,J_z)) - \vec{\xi}_{n+2}^T D^T_{n+1}C_{n+1} \\
	\quad &\text{3) } \text{Output: } V = V(x,y,z) \approx \xi_0 Y_0
}
where at each iteration, $\vec{\xi}_n$ is a vector of matrices\footnote{It may help to think of this symbolically. At each iteration, the entries of $\vec{\xi}_n$ are polynomials in $J_x$, $J_y$ and $J_z$. On the final step, $\vec{\xi}_{0} = \xi_{0}$ is hence some polynomial expression in $J_x$, $J_y$ and $J_z$, which we can then evaluate by inserting the Jacobi matrices.} \bstodo{Added footnote to help explain better what the notation here is meaning.}(note that we are abusing notation here a bit, however it is the simplest way to present the algorithm without introducing yet more matrices!).

@@

The Jacobi matrices are then defined according to the matrices \nomenclature[JT]{$\tangentjacobi_x, \tangentjacobi_y, \tangentjacobi_z$}{The Jacobi matrices for the tangent bundle VSH basis} $\tangentjacobi_x, \tangentjacobi_y, \tangentjacobi_z$ satisfying \bstodo{Added the notation for $\tangentjacobi_x$, etc. to the index.}

@@

The framework we have outlined for the spherical harmonics and the vector spherical harmonics allows us to easily and explicitly derive sparse operator matrices for partial differential operators. Of course, not all PDEs can be sparsely represented, however we can do this for key operators such as the spherical gradient, divergence and Laplace--Beltrami, as well as other operators found in examples such as the linearised shallow water equations.\bstodo{Added the comment that not all PDEs operators can be sparsely represented.} These operators will take coefficients of a function's expansion in either the SH or VSH basis, to coefficients in either basis.

@@

\begin{theorem}
	The operator matrices defined in \bsrefdef{def:SHoperators} are diagonal, and are given by: \bstodo{Tried to make clearer what $\calG$ is here.}
\bseqn{
	\calG &= I_{(N+1)^2}, \quad
	\tilde \calG = 
		\begin{pmatrix}
			1 & & & & \\
			& 0 & & & \\
			& & 1 & & & & \\
			& & & 0 & & & \\
			& & & & \ddots & & \\
			& & & & & 1 & \\
			& & & & & & 0
		\end{pmatrix}  \in \R^{2(N+1)^2 \times 2(N+1)^2}, \\
	\calD &\equiv \calL := 
		\begin{pmatrix}
			\calL_0 & & \\
			& \ddots & \\
			& & \calL_N
		\end{pmatrix} \in \R^{(N+1)^2 \times (N+1)^2}, \\
	\tilde \calD &= 
		\begin{pmatrix}
			\tilde \calD_0 & & \\
			& \ddots & \\
			& & \tilde \calD_N
		\end{pmatrix} \in \R^{2(N+1)^2 \times 2(N+1)^2},
}
where
\bseqn{
	\calL_l := -l(l+1) \: I_{2l+1}, \quad
	\tilde \calD_l := \begin{pmatrix}
			-l(l+1) & & & & \\
			& 0 & & & \\
			& & \ddots & & \\
			& & & -l(l+1) & \\
			& & & & 0
		\end{pmatrix} \in \R^{2(2l+1) \times 2(2l+1)}.
}.
\end{theorem}
